---
title: "Part 6: Model Fitting with a Single Available GWIS (MERLIN-MO and MERLIN-ME)"
date: "\\small \\today"
output:
  html_document:
    toc: yes
    toc_depth: '3'
    df_print: paged
editor_options:
  markdown:
    wrap: 72
vignette: >
  %\VignetteIndexEntry{Part 6: Model Fitting with a Single Available GWIS}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(encoding = "UTF-8")
```

## Introduction to the Single-GWIS Extensions

GWIS summary statistics may be unavailable for either the exposure or the
outcome. MERLIN therefore provides two reduced-data extensions:

- **MERLIN-MO** is used when the **outcome GWIS is missing**. In the R
  package, this model is selected using `model = "MO"`, where
  $\widehat{\Gamma}_3$ denotes the outcome GWIS association.

- **MERLIN-ME** is used when the **exposure GWIS is missing**. In the R
  package, this model is selected using `model = "ME"`, where
  $\widehat{\gamma}_3$ denotes the exposure GWIS association.

The distinction between uppercase $\Gamma$ and lowercase $\gamma$ is
important: uppercase quantities refer to the outcome, whereas lowercase
quantities refer to the exposure.

| Method | Exposure GWAS | Exposure GWIS | Outcome GWAS | Outcome GWIS | R model option |
|:--|:--:|:--:|:--:|:--:|:--|
| MERLIN-MO | Required | Required | Required | Not used | `"MO"` |
| MERLIN-ME | Required | Not used | Required | Required | `"ME"` |

This tutorial follows the simulation framework introduced in Part 3 and
demonstrates these two settings separately. Each analysis uses two GWAS inputs
and the single GWIS input. 

## Step-by-Step Simulation

First, we load the required packages and set a random seed for
reproducibility.

```{r}
library(MERLIN)
library(mvtnorm)
set.seed(2027)
```

**Generating Genotypes and the Modifier**

As in Part 3, we simulate $m = 100$ independent SNPs for 160,000 individuals
and divide them equally into an exposure cohort and an outcome cohort. The
modifier $E$ is generated from a standard normal distribution.

```{r, eval=FALSE}
n_exp <- 80000
n_out <- 80000
m <- 100

# True causal parameters
b1 <- 0
b4 <- 0.3

# Generate genotype matrix G
maf <- runif(m, 0.05, 0.5)
G <- matrix(
  rbinom((n_exp + n_out) * m, 2,
         rep(maf, each = n_exp + n_out)),
  nrow = n_exp + n_out,
  ncol = m
)
G <- scale(G, center = TRUE, scale = FALSE)

# Generate modifier E
E_exp <- sample(rep(c(-1, 1), each = n_exp / 2))
E_out <- sample(rep(c(-1, 1), each = n_out / 2))
E <- c(E_exp, E_out)
```

**Simulating Genetic Effects**

The main genetic effects ($\gamma_1$) and SNP-by-modifier interaction effects
($\gamma_3$) are simulated from a bivariate normal distribution. Their
variances are controlled by `h_g1` and `h_g3`, and `cor_g1g3` determines the
correlation between the two sets of genetic effects.

```{r, eval=FALSE}
h_g1 <- 0.3
h_g3 <- 0.1
h_b <- 0.05
cor_g1g3 <- 0

sigma2g1 <- h_g1 / m
sigma2g3 <- h_g3 / m
sigma2b  <- h_b / m

cov_matrix <- matrix(c(sigma2g1, 
                      cor_g1g3 * sqrt(sigma2g1 * sigma2g3),
                      cor_g1g3 * sqrt(sigma2g1 * sigma2g3),
                      sigma2g3), ncol = 2)

gamma1_3 <- rmvnorm(m, mean = c(0, 0), sigma = cov_matrix)
gamma_1x <- gamma1_3[, 1]
gamma_3x <- gamma1_3[, 2]
beta_2   <- rnorm(m, 0, sqrt(sigma2b))
```

**Generating Exposure and Outcome Phenotypes**

We use the same two-sample design as in Part 3. The residual errors for $X$
and $Y$ are correlated to represent unmeasured confounding, while the exposure
and outcome summary statistics are calculated in non-overlapping cohorts.

```{r, eval=FALSE}
# Construct the GxE term
GE <- G * E

# Correlation between the residual errors
rhoxy <- 0.6

var_noise_x <- 1 - h_g1 - h_g3
var_noise_y <- 1 - b1*b1 - h_b - b4*b4

# Construct the covariance matrix for the bivariate normal distribution
cov_xy <- rhoxy * sqrt(var_noise_x * var_noise_y)
sigma_noise <- matrix(c(var_noise_x, cov_xy,
                        cov_xy, var_noise_y), ncol = 2)

# Generate correlated noise for X and Y
noise_xy <- rmvnorm(n_exp + n_out, mean = c(0, 0), sigma = sigma_noise)
noise_x <- noise_xy[, 1]
noise_y <- noise_xy[, 2]

# Generate exposure and outcome
X <- G %*% gamma_1x + GE %*% gamma_3x + noise_x
Y <- X * b1 + G %*% beta_2 + X * E * b4 + noise_y

# Split into non-overlapping exposure and outcome cohorts
idx_exp <- seq_len(n_exp)
idx_out <- n_exp + seq_len(n_out)

exp_pheno <- X[idx_exp]
exp_E <- E[idx_exp]

out_pheno <- Y[idx_out]
out_E <- E[idx_out]
```


## Obtaining Summary Statistics

We use the same helper function as in Part 3 to obtain marginal GWAS and GWIS
associations. For a GWIS, the coefficient and standard error of the
$G_j \times E$ term are extracted from a single-variant interaction model.

```{r, eval=FALSE}
get_sumstats <- function(G, pheno, interaction = FALSE, E = NULL) {
  betas <- numeric(ncol(G))
  ses <- numeric(ncol(G))

  for (i in seq_len(ncol(G))) {
    if (interaction) {
      model <- lm(pheno ~ G[, i] + E + G[, i]:E)
      betas[i] <- coef(model)[4]
      ses[i] <- summary(model)$coefficients[4, 2]
    } else {
      model <- lm(pheno ~ G[, i])
      betas[i] <- coef(model)[2]
      ses[i] <- summary(model)$coefficients[2, 2]
    }
  }

  list(beta = betas, se = ses)
}

# GWAS inputs shared by MERLIN-MO and MERLIN-ME
exp_gwas_sum <- get_sumstats(
  G[idx_exp, ], exp_pheno
)
out_gwas_sum <- get_sumstats(
  G[idx_out, ], out_pheno
)
```

**Available GWIS for MERLIN-MO**

MERLIN-MO assumes that the exposure GWIS is available but the outcome GWIS is
unavailable. We therefore calculate only the exposure GWIS needed for this
analysis.

```{r, eval=FALSE}
exp_gwis_sum <- get_sumstats(
  G[idx_exp, ], exp_pheno,
  interaction = TRUE, E = exp_E
)
```

**Available GWIS for MERLIN-ME**

MERLIN-ME assumes that the outcome GWIS is available but the exposure GWIS is
unavailable. We therefore calculate only the outcome GWIS needed for this
analysis.

```{r, eval=FALSE}
out_gwis_sum <- get_sumstats(
  G[idx_out, ], out_pheno,
  interaction = TRUE, E = out_E
)
```

The two GWIS objects above belong to different demonstrations. `out_gwis_sum`
is never used by MERLIN-MO, and `exp_gwis_sum` is never used by MERLIN-ME.


## Instrument Selection

For this toy example, SNPs are independent and we use $P < 5e-8$ as the
instrument-selection threshold. In real applications, genome-wide thresholds
and LD clumping should be used as described in Part 2.

```{r, eval=FALSE}
p_threshold <- 5e-8

# Exposure-GWAS instruments
pvals_gwas <- 2 * pnorm(-abs(exp_gwas_sum$beta / exp_gwas_sum$se))
iv_gwas <- which(pvals_gwas < p_threshold)

# Exposure-GWIS instruments
pvals_gwis <- 2 * pnorm(-abs(exp_gwis_sum$beta / exp_gwis_sum$se))
iv_gwis <- which(pvals_gwis < p_threshold)

# MERLIN-MO uses the union of exposure-GWAS and exposure-GWIS instruments.
iv_mo <- union(iv_gwas, iv_gwis)
R_mo <- diag(length(iv_mo))

# MERLIN-ME cannot use exposure-GWIS information. Its instruments are
# therefore selected from the exposure GWAS alone.
iv_me <- iv_gwas
R_me <- diag(length(iv_me))
```

## Model Fitting

### MERLIN-MO: Outcome GWIS unavailable

When outcome GWIS summary statistics are unavailable, $\widehat{\Gamma}_3$ and its standard error are omitted. MERLIN-MO uses the exposure GWAS, exposure GWIS, and outcome GWAS summary statistics to jointly estimate the average causal effect $\beta_1$ and the interaction effect $\beta_4$ within the reduced-data model. Although the outcome GWIS is unavailable, uncertainty in both causal parameters is retained during model fitting.

```{r, eval=FALSE}
# Fit MERLIN-MO using the exposure GWAS, exposure GWIS, and outcome GWAS.

fit_mo <- MERLIN(
  gammah1 = exp_gwas_sum$beta[iv_mo],
  gammah3 = exp_gwis_sum$beta[iv_mo],
  Gammah1 = out_gwas_sum$beta[iv_mo],
  se1 = exp_gwas_sum$se[iv_mo],
  se2 = exp_gwis_sum$se[iv_mo],
  se3 = out_gwas_sum$se[iv_mo],
  R = R_mo,
  rho_1 = 0,
  model = "MO",
  seed = 2027
)
str(fit_mo)
```

```text
List of 6
 $ Beta1.hat : num -0.021
 $ Beta1.se  : num 0.0453
 $ Beta1.pval: num 0.643
 $ Beta4.hat : num 0.291
 $ Beta4.se  : num 0.0702
 $ Beta4.pval: num 3.46e-05
 ```

### MERLIN-ME: Exposure GWIS unavailable

When exposure GWIS summary statistics are unavailable, $\widehat{\gamma}_3$
and its standard error are omitted. MERLIN-ME estimates $\beta_1$ by
MR-Egger regression of the outcome-GWAS associations on the exposure-GWAS
associations. It estimates $\beta_4$ by a second MR-Egger regression of the
outcome-GWIS associations on the exposure-GWAS associations.

```{r, eval=FALSE}
fit_me <- MERLIN(
  gammah1 = exp_gwas_sum$beta[iv_me],
  Gammah1 = out_gwas_sum$beta[iv_me],
  Gammah3 = out_gwis_sum$beta[iv_me],
  se1 = exp_gwas_sum$se[iv_me],
  se3 = out_gwas_sum$se[iv_me],
  se4 = out_gwis_sum$se[iv_me],
  R = R_me,
  model = "ME"
)
str(fit_me)
```

```text
List of 6
 $ Beta1.hat : num 0.122
 $ Beta1.se  : num 0.079
 $ Beta1.pval: num 0.125
 $ Beta4.hat : num 0.291
 $ Beta4.se  : num 0.0198
 $ Beta4.pval: num 1.37e-24
```

Neither `gammah3` nor `se2` is supplied, and the exposure GWIS is not used to
select `iv_me`.
